Executive Constraints and Civil Conflict Onset

### 2020-08-31 paper.Rmd
Final edits

---
title: Executive Constraints and Civil Conflict Onset
author:
    name: J.L.A. Krusell
    affiliation: Department of Peace and Conflict Research
    supervisor: Kristine Eck
    degree: "Master's Thesis"
date: "Spring 2020"
titlelogo: |
    ```{=latex}
    assets/UU_logo_CMYK-eps-converted-to.pdf
    ```
wordcount: 15000
suppress-bibliography: false
bibliography: library.bib
csl: assets/chicago-author-date.csl
link-citations: true
documentclass: scrreprt
classoption: paper=letter
geometry:
    - top=30mm
    - heightrounded
fontsize: 12pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
    - \usepackage[font={small}]{caption}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--template", "assets/thesis_template.latex",
                      "--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate",
                      "--lua-filter", "utils/text_refs/text_refs.lua"]
    html_document:
        smart: true
        toc: true
abstract: |
    **Abstract**: Do institutional constraints on a regime's executive decrease the
    likelihood of civil conflict onset? An unconstrained executive is free
    to loot state resources, undermining the state's capacity to
    effectively deal with nascent rebellions, and may be more likely to
    engage in violent repression, especially in the form of indiscriminate
    violence. This can encourage political grievances, lead to a loss of
    legitimacy for the regime, and provide the opportunity for would be rebel
    groups to attract new members by offering protection against
    government violence. Furthermore, the lack of guaranteed checks to
    executive power may incentivise actors to avoid bargained solutions if
    they fear future extralegal retributions. Taken together, where there
    is a lack of effective constraints against the executive there may be
    an increased risk for the onset of civil conflict. To test this
    proposition I employ a bayesian latent variable model which estimates
    executive constraint as a latent factor derived from several manifest
    variables and relates it to a binary measure of civil conflict onset
    through a logistic regression.  The primary finding is that there is a
    negative relationship such that the predicted probability for civil
    conflict is lower where there exists higher levels of executive
    constraints. However, this is conditional on the level of GDP per
    capita; for low-income country-years the relationship between
    executive constraints and civil conflict onset is indeterminate
    possibly because it is easier to recruit and mobilize fighters
    regardless of the overall level of executive constraints. The issue of
    poor model fit means that the presented results should be considered
    tentative at best.
---

@@

variable_names <- c("rgdpepc" = "GDPpc (ln)",
                    "rgdpepc_gro" = "GDPpc Growth",
                    "cinc" = "CINC",
                    "discrimpop" = "Discriminated Pop.",
                    "pop_density" = "Pop. Density (ln)",
                    "meanelev" = "Mean Elevation (ln)",
                    "rlvt_groups_count" = "Relevant Groups",
                    "neighbour_conflict" = "Neighbouring Conflict",
                    "major_ongoing" = "High Int. Conflict",
                    "independence" = "Independence",
                    "exec_constraints" = "Exec. Constraints",
                    "exec * rgdpepc" = "Exec. Constraints * GDPpc")

@@

The principle finding of this paper is that there is tentative
evidence that executive constraints do matter and that there is a
negative relationship between the latent level of constraints on the
executive and the risk of civil conflict onset. However, this
relationship largely disappears for low-income country-years when an
interaction term between executive constraints and gross domestic
product (GDP) per capita is included. If the primary theoretical
mechanism involves the impact of executive constraints on rebel
mobilization --- an unchecked executive may be more likely to engage
in indiscriminate violence which could lead to a loss of legitimacy as
well as the ability of rebel groups to offer local security guarantees
to attract recruits --- then the incentives for violence in low-income
country-years due to say economic factors may be so great that it does
not matter how constrained the executive may or may not be.

@@

In what follows, a deeper justification for why political institutions
are relevant to the study of civil conflict together with an overview
of related empirical literature will be presented. The theoretical
mechanisms linking executive constraints to conflict onset will then
be discussed followed by an empirical analysis using a bayesian latent
variable model to estimate executive constraints from the basis of
several observed manifest variables. This latent measure will then be
used to predict civil conflict onset using a cross-country time series
logistic regression.

@@

(ref:onsets) Proportions of countries per year experiencing at least
    one ongoing civil conflict and at least one new onset.

@@

ggplot(x, aes(theta, point, group = group, col = factor)) +
    geom_line(alpha = 0.3) +
    geom_point(alpha = 0.8) +
    geom_line(data = y, aes(theta, yhat, group = NA), colour = "black", alpha = 1) +
    theme_tufte() +
    scale_color_colorblind() +
    theme(legend.title = element_blank()) +
    ylab(target) +
    xlab("Theta")

@@

#### Random Effects

@@

From the listed plots it is clear that for medium to high income
country-years the overall effect of executive constraints when
interacted with GDP per capita on the predicted probability of
conflict onset is negative --- that is to say a more constrained
executive is associated with a decreased risk of civil conflict onset.
For example, for country-years with a relatively high level of GDP per
capita, going from the worst to best in executive constraints
decreases the predicted probability from [1.5, 4.5] down to below
0.5. This effect diminishes, however, for lower values of GDP per
capita to the point that executive constraints are practically
indeterminate to the risk of conflict onset for country-years situated
around the 10\% quantile for GDP per capita due to the high degree of
posterior uncertainty. In the plot depicting low GDP per capita, there
is, nonetheless, a slight increase in the median predicted
probability, which begs the question: for GDP per capita values lower
than the 10\% quantile is there a distinguishable reversal in the
relationship between executive constraints and civil conflict such
that there is actually a *positive* increase in the predicted
probability for higher levels of executive constraints?

@@

ggplot(df, aes(original, median)) +
    geom_line() +
    geom_ribbon(aes(ymin = codelow, ymax = codehigh), alpha = 0.2) +
    theme_tufte() +
    xlab("GDP per Capita")
```

@@

(ref:confusion) Confusion matrix with a 0.168 cutoff for the median
    posterior predicted probabilities.

```{r confusion}
m <- table(p_median > .168, final.df$lepisode_onset)

@@

To assess how well the presented model fits the observed data, I first
consider the predicted probabilities of conflict onset by plotting the
first 50 posterior draws for two groups, country-years that actually
experienced an onset of civil conflict and country-years that did
not. On average, the predicted probabilities for the actual onset
country-years are higher albeit still extremely low. This can be seen
in the summary statistics reported in table \ref{tab:psum} for the
posterior medians of the predicted probabilities. For onset
country-years the interquartile interval is only `r df[1, "1st Qu."]`
to `r df[1, "3rd Qu."]` with a max of `r df[1, "Max."]`. To better
compare the predicted probabilities to the actual data, table
\ref{tab:confusion} presents a confusion matrix with a 0.168 cutoff,
which corresponds to the third quartile in the summary table. `r tpr`
percent of onsets and `r tnr` percent of non-onsets are correctly
predicted. The false positive rate is `r fpr` percent and the false
negative rate is `r fnr` percent.

@@

(ref:confusion-major) Confusion matrix with a 0.168 cutoff for the
    median posterior predicted probabilities from the single year high
    intensity model run.

@@

m <- table(p_median > .168, major$final.df$lmajor_onset)
rownames(m) <- c("Predicted: No Onset", "Predicted: Onset")
colnames(m) <- c("Actual: No Onset", "Actual: Onset")

@@

Table \ref{tab:confusion-major} presents the confusion matrix for the
single year high intensity model run using the median posterior
predicted probabilities. A cutoff of 0.168 was reused for comparative
purposes to the original results. In this case, the model is unable to
predict even a single major conflict onset with the highest median
posterior predicted probability for an actual onset being only
`r fmt(max(p_median[major$final.df$lmajor_onset == 1]))`. Furthermore,
although the ROC AUC is `r major_auc`, this is again due to the major
class imbalance between positive versus negative outcomes in the
dataset. A precision-recall AUC of `r major_auc_pr` demonstrates the
poor performance of the model.

(ref:confusion-cumulative) Confusion matrix with a 0.168 cutoff for
    the median posterior predicted probabilities from the cumulative
    high intensity model run.

@@

m <- table(p_median > .168, cumulative$final.df$lcum_onset)
rownames(m) <- c("Predicted: No Onset", "Predicted: Onset")
colnames(m) <- c("Actual: No Onset", "Actual: Onset")

>>>>

---
title: Executive Constraints and Civil Conflict Onset
author:
    name: J.L.A. Krusell
    affiliation: Department of Peace and Conflict Research
    supervisor: Kristine Eck
    degree: "Master's Thesis"
date: "Spring 2020"
titlelogo: |
    ```{=latex}
    assets/UU_logo_CMYK-eps-converted-to.pdf
    ```
wordcount: 15000
suppress-bibliography: false
bibliography: library.bib
csl: assets/chicago-author-date.csl
link-citations: true
documentclass: scrreprt
classoption: paper=letter
geometry:
    - top=30mm
    - heightrounded
fontsize: 12pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
    - \usepackage[font={small}]{caption}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--template", "assets/thesis_template.latex",
                      "--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate",
                      "--lua-filter", "utils/text_refs/text_refs.lua"]
    html_document:
        smart: true
        toc: true
abstract: |
    **Abstract**: Do institutional constraints on a regime's executive
    decrease the likelihood of civil conflict onset? An unconstrained
    executive is free to loot state resources undermining the state's
    capacity to effectively deal with nascent rebellions, and may be
    more likely to engage in violent repression, especially in the
    form of indiscriminate violence. This can encourage political
    grievances, lead to a loss of legitimacy for the regime, and
    provide the opportunity for would be rebel groups to attract new
    members by offering protection against government
    violence. Furthermore, the lack of guaranteed checks to executive
    power may incentivise actors to avoid bargained solutions if they
    fear future extralegal retributions. Taken together, where there
    is a lack of effective constraints against the executive there may
    be an increased risk for the onset of civil conflict. To test this
    proposition I employ a bayesian latent variable model which
    estimates executive constraint as a latent factor derived from
    several manifest variables and relate it to a binary measure of
    civil conflict onset through a logistic regression.  The primary
    finding is that there is a negative relationship such that the
    predicted probability for civil conflict is lower where there
    exists higher levels of executive constraints. However, this is
    conditional on the level of GDP per capita; for low-income
    country-years the relationship between executive constraints and
    civil conflict onset is indeterminate possibly because it is
    easier to recruit and mobilize fighters in such settings
    regardless of the overall level of executive constraints. The
    model, however, is a poor fit to the data meaning that the
    presented results should be considered tentative at
    best. Nonetheless, this study helps to further the work on
    examining specific political institutions as potential risk
    factors for the onset of civil conflict.
---

@@

variable_names <- c("rgdpepc" = "GDPpc (ln)",
                    "rgdpepc_gro" = "GDPpc Growth",
                    "cinc" = "CINC",
                    "discrimpop" = "Discriminated Pop.",
                    "pop_density" = "Pop. Density (ln)",
                    "meanelev" = "Mean Elevation (ln)",
                    "rlvt_groups_count" = "Relevant Groups",
                    "neighbour_conflict" = "Neighbouring Conflict",
                    "major_ongoing" = "Major Conflict",
                    "independence" = "Independence",
                    "exec_constraints" = "Exec. Constraints",
                    "exec * rgdpepc" = "Exec. Constraints * GDPpc")

@@

To study the potential link between executive constraints and civil
conflict onset I use a bayesian latent variable model to estimate
executive constraints as a latent factor drawn from the basis of
several observed manifest variables. I then relate this measure to a
binary indicator for civil conflict onset using a cross-country time
series logistic regression.

The principle finding is that there is tentative evidence that
executive constraints do matter and that there is a negative
relationship between the latent level of constraints on the executive
and the risk of civil conflict onset. However, this relationship
largely disappears for low-income country-years when an interaction
term between executive constraints and gross domestic product (GDP)
per capita is included. If the primary theoretical mechanism involves
the impact of executive constraints on rebel mobilization --- an
unchecked executive may be more likely to engage in indiscriminate
violence which could lead to a loss of legitimacy as well as the
ability of rebel groups to offer local security guarantees to attract
recruits --- then the incentives for violence in low-income
country-years due to say economic factors may be so great that it does
not matter how constrained the executive may or may not be.

@@

In what follows, a deeper justification for why political institutions
are relevant to the study of civil conflict together with an overview
of related empirical literature will be presented. The theoretical
mechanisms linking executive constraints to conflict onset will then
be discussed followed by a presentation of the paper's research design
and empirical analysis, including both the latent variable measurement
model for executive constraints and the cross-country time series
logistic regression predicting civil conflict onset. The final section
provides conclusions and possible suggestions for future research.

@@

(ref:onsets) Proportions of countries experiencing at least one
    ongoing civil conflict and at least one new onset per year.

@@

ggplot(x, aes(theta, point, group = group, col = factor)) +
    geom_line(alpha = 0.3) +
    geom_point(alpha = 0.8) +
    geom_line(data = y, aes(theta, yhat, group = NA), colour = "black", alpha = 1) +
    theme_tufte() +
    scale_color_colorblind() +
    theme(legend.title = element_blank()) +
    ylab(target) +
    xlab("Executive Constraints")

@@

#### Varying Intercepts

@@

From the listed plots it is clear that for medium to high income
country-years the overall effect of executive constraints when
interacted with GDP per capita on the predicted probability of
conflict onset is negative --- that is to say a more constrained
executive is associated with a decreased risk of civil conflict onset.
For example, for country-years with a relatively high level of GDP per
capita, going from the worst to best in executive constraints
decreases the predicted probability from [1.0, 4.5] down to below
0.5. This effect diminishes, however, for lower values of GDP per
capita to the point that executive constraints are practically
indeterminate to the risk of conflict onset for country-years situated
around the 10\% quantile for GDP per capita due to the high degree of
posterior uncertainty. In the plot depicting low GDP per capita, there
is, nonetheless, a slight increase in the median predicted
probability, which begs the question: for GDP per capita values lower
than the 10\% quantile is there a distinguishable reversal in the
relationship between executive constraints and civil conflict such
that there is actually a *positive* increase in the predicted
probability for higher levels of executive constraints?

@@

ggplot(df, aes(original, median)) +
    geom_line() +
    geom_ribbon(aes(ymin = codelow, ymax = codehigh), alpha = 0.2) +
    theme_tufte() +
    xlab("GDP per Capita") +
    theme(axis.title.y = element_blank())

@@

(ref:confusion) Confusion matrix with a 0.17 cutoff for the median
    posterior predicted probabilities.

```{r confusion}
m <- table(p_median > .17, final.df$lepisode_onset)

@@

To assess how well the presented model fits the observed data, I first
consider the predicted probabilities of conflict onset by plotting the
first 50 posterior draws for two groups, country-years that actually
experienced an onset of civil conflict and country-years that did
not. On average, the predicted probabilities for the actual onset
country-years are higher albeit still extremely low. This can be seen
in the summary statistics reported in table \ref{tab:psum} for the
posterior medians of the predicted probabilities. For onset
country-years the interquartile interval is only `r df[1, "1st Qu."]`
to `r df[1, "3rd Qu."]` with a max of `r df[1, "Max."]`. To better
compare the predicted probabilities to the actual data, table
\ref{tab:confusion} presents a confusion matrix with a 0.17 cutoff,
which corresponds to the third quartile in the summary table. `r tpr`
percent of onsets and `r tnr` percent of non-onsets are correctly
predicted. The false positive rate is `r fpr` percent and the false
negative rate is `r fnr` percent.

@@

(ref:confusion-major) Confusion matrix with a 0.17 cutoff for the
    median posterior predicted probabilities from the single year high
    intensity model run.

@@

m <- table(p_median > .17, major$final.df$lmajor_onset)
rownames(m) <- c("Predicted: No Onset", "Predicted: Onset")
colnames(m) <- c("Actual: No Onset", "Actual: Onset")

@@

Table \ref{tab:confusion-major} presents the confusion matrix for the
single year high intensity model run using the median posterior
predicted probabilities. A cutoff of 0.17 was reused for comparative
purposes to the original results. In this case, the model is able to
only correctly predict a single major conflict onset with the highest
median posterior predicted probability for an actual onset being only
`r fmt(max(p_median[major$final.df$lmajor_onset == 1]))`. Furthermore,
although the ROC AUC is `r major_auc`, this is again due to the major
class imbalance between positive versus negative outcomes in the
dataset. A precision-recall AUC of `r major_auc_pr` demonstrates the
poor performance of the model.

(ref:confusion-cumulative) Confusion matrix with a 0.17 cutoff for
    the median posterior predicted probabilities from the cumulative
    high intensity model run.

@@

m <- table(p_median > .17, cumulative$final.df$lcum_onset)
rownames(m) <- c("Predicted: No Onset", "Predicted: Onset")
colnames(m) <- c("Actual: No Onset", "Actual: Onset")
